---
title: "Building queries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building queries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

Queries are trees of clauses combined under AND/OR operators. Build leaves
with `createClause()`, combine them with `buildClauseGroup()`, nest groups
to express arbitrary Boolean logic.

## Clause types

```{r}
library(picsure)

# Categorical FILTER
sex_filter <- picsure::createClause(
  "\\phs000001\\pht000001\\phv00000001\\sex\\",
  type       = "FILTER",
  categories = list("male")
)

# Continuous FILTER
age_filter <- picsure::createClause(
  "\\phs000001\\pht000001\\phv00000005\\age\\",
  type = "FILTER", min = 40
)

# SELECT
bmi_select <- picsure::createClause(
  "\\phs000001\\pht000001\\phv00000020\\bmi\\",
  type = "SELECT"
)

# REQUIRE
smoking_require <- picsure::createClause(
  "\\phs000001\\pht000001\\phv00000030\\smoking\\",
  type = "REQUIRE"
)

# ANYRECORD
visits_anyrecord <- picsure::createClause(
  "\\phs000001\\pht000002\\phv00000100\\visit_count\\",
  type = "ANYRECORD"
)
```

## Combine with AND

```{r}
query <- picsure::buildClauseGroup(
  list(sex_filter, age_filter, bmi_select),
  operator = "AND"
)
```

## Nested groups for AND-of-ORs

A cohort of males aged 40+ who have either COPD or asthma:

```{r}
copd_filter <- picsure::createClause(
  "\\phs000001\\pht000002\\phv00000050\\copd\\",
  type = "FILTER", categories = list("Yes")
)
asthma_filter <- picsure::createClause(
  "\\phs000001\\pht000002\\phv00000051\\asthma\\",
  type = "FILTER", categories = list("Yes")
)

copd_or_asthma <- picsure::buildClauseGroup(
  list(copd_filter, asthma_filter),
  operator = "OR"
)

full_query <- picsure::buildClauseGroup(
  list(sex_filter, age_filter, copd_or_asthma),
  operator = "AND"
)
```

## Case-insensitive enums

Both `type` (clause type) and `operator` (group operator) accept
case-insensitive strings: `"filter"`, `"Filter"`, `"FILTER"` are all
equivalent.
