---
title: "Building queries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building queries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

Queries are trees of clauses combined under AND/OR operators. Build leaves
with `createSubQuery()`, combine them with `buildQuery()`, nest groups
to express arbitrary Boolean logic.

## Clause types

```{r}
library(picsure)

# Categorical FILTER
sex_filter <- picsure::createSubQuery(
  "\\phs000001\\pht000001\\phv00000001\\sex\\",
  type       = "FILTER",
  categories = list("male")
)

# Continuous FILTER
age_filter <- picsure::createSubQuery(
  "\\phs000001\\pht000001\\phv00000005\\age\\",
  type = "FILTER", min = 40
)

# SELECT
bmi_select <- picsure::createSubQuery(
  "\\phs000001\\pht000001\\phv00000020\\bmi\\",
  type = "SELECT"
)

# REQUIRE
smoking_require <- picsure::createSubQuery(
  "\\phs000001\\pht000001\\phv00000030\\smoking\\",
  type = "REQUIRE"
)

# ANYRECORD
visits_anyrecord <- picsure::createSubQuery(
  "\\phs000001\\pht000002\\phv00000100\\visit_count\\",
  type = "ANYRECORD"
)
```

You can also pass an enum member instead of a string for editor autocomplete:

```{r enum-member-form, eval=FALSE}
# equivalent to type = "FILTER":
sex_filter <- picsure::createSubQuery(
  "\\phs000001\\pht000001\\phv00000001\\sex\\",
  type       = picsure::PhenotypicFilterType$FILTER,
  categories = list("male")
)
```

The same applies to `buildQuery(operator = picsure::GroupOperator$AND)` and `runQuery(type = picsure::QueryType$COUNT)`. Strings remain valid for backwards compatibility; the enum members are new in this release.

## Combine with AND

```{r}
query <- picsure::buildQuery(
  list(sex_filter, age_filter, bmi_select),
  operator = "AND"
)
```

## Nested groups for AND-of-ORs

A cohort of males aged 40+ who have either COPD or asthma:

```{r}
copd_filter <- picsure::createSubQuery(
  "\\phs000001\\pht000002\\phv00000050\\copd\\",
  type = "FILTER", categories = list("Yes")
)
asthma_filter <- picsure::createSubQuery(
  "\\phs000001\\pht000002\\phv00000051\\asthma\\",
  type = "FILTER", categories = list("Yes")
)

copd_or_asthma <- picsure::buildQuery(
  list(copd_filter, asthma_filter),
  operator = "OR"
)

full_query <- picsure::buildQuery(
  list(sex_filter, age_filter, copd_or_asthma),
  operator = "AND"
)
```

## Editing an existing query

`removeSubQuery()` and `replaceClause()` let you edit a query tree
without rebuilding it from scratch. Both return a **new** handle — the
input is not mutated. Matching is structural: any nested clause or
clause-group that is equal by value to `target` is matched.

```{r}
# Drop the copd_or_asthma sub-tree:
without_lung <- picsure::removeSubQuery(full_query, copd_or_asthma)

# Swap "Yes" for a more specific COPD answer set:
copd_specific <- picsure::createSubQuery(
  "\\phs000001\\pht000002\\phv00000050\\copd\\",
  type = "FILTER",
  categories = list("Yes, severe", "Yes, very severe")
)
refined <- picsure::replaceClause(full_query, copd_filter, copd_specific)
```

`removeSubQuery()` drops `ClauseGroup`s that are emptied by a removal,
so you don't end up with orphan operators. It errors if the removal
would empty the entire tree — build a fresh query instead in that case.

## Save a query by name

On authorized deployments, persist a query to your user profile under
a display name with `saveQueryByName()`. It returns the PIC-SURE query
ID, which you can later pass to `loadQueryByID()` or `runQueryByID()`:

```{r}
qid <- picsure::saveQueryByName(bdc, full_query, "Cohort 2026-Q2")

# Re-point an existing name at a refined query:
qid <- picsure::saveQueryByName(bdc, refined, "Cohort 2026-Q2", overwrite = TRUE)
```

Allowed characters in `name`: letters, digits, spaces, and
``- _ \ / ? + = [ ] . ( ) : " '`` (max 255 chars; validated by the
Python adapter against the backend `NamedDataset` pattern). Not
supported on open-access platforms.

## Case-insensitive enums

Both `type` (clause type) and `operator` (group operator) accept
case-insensitive strings: `"filter"`, `"Filter"`, `"FILTER"` are all
equivalent.
