---
title: "Building queries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building queries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

Queries are trees of clauses combined under AND/OR operators. Build leaves
with `buildClause()`, combine them with `buildClauseGroup()`, nest groups to
express arbitrary Boolean logic, and — when you want to choose which concept
paths come back in the output — assemble the whole thing with `buildQuery()`.

## Clause types

```{r}
library(picsure)

# Categorical FILTER
sex_filter <- picsure::buildClause(
  "\\phs000001\\pht000001\\phv00000001\\sex\\",
  type       = "FILTER",
  categories = list("male")
)

# Continuous FILTER
age_filter <- picsure::buildClause(
  "\\phs000001\\pht000001\\phv00000005\\age\\",
  type = "FILTER", min = 40
)

# REQUIRE
smoking_require <- picsure::buildClause(
  "\\phs000001\\pht000001\\phv00000030\\smoking\\",
  type = "REQUIRE"
)

# ANYRECORD
visits_anyrecord <- picsure::buildClause(
  "\\phs000001\\pht000002\\phv00000100\\visit_count\\",
  type = "ANYRECORD"
)
```

To **include a concept path in the output without filtering** (what the old
`"SELECT"` clause type did), pass it to `buildQuery(includeConcepts = ...)` —
see "Choosing output concepts" below. Output columns are no longer a clause
type.

You can also pass an enum member instead of a string for editor autocomplete:

```{r enum-member-form, eval=FALSE}
# equivalent to type = "FILTER":
sex_filter <- picsure::buildClause(
  "\\phs000001\\pht000001\\phv00000001\\sex\\",
  type       = picsure::PhenotypicFilterType$FILTER,
  categories = list("male")
)
```

The same applies to `buildClauseGroup(operator = picsure::GroupOperator$AND)` and `runQuery(type = picsure::QueryType$COUNT)`. Strings remain valid for backwards compatibility; the enum members are new in this release.

## Combine with AND

```{r}
filters <- picsure::buildClauseGroup(
  list(sex_filter, age_filter),
  operator = "AND"
)
```

## Nested groups for AND-of-ORs

A cohort of males aged 40+ who have either COPD or asthma:

```{r}
copd_filter <- picsure::buildClause(
  "\\phs000001\\pht000002\\phv00000050\\copd\\",
  type = "FILTER", categories = list("Yes")
)
asthma_filter <- picsure::buildClause(
  "\\phs000001\\pht000002\\phv00000051\\asthma\\",
  type = "FILTER", categories = list("Yes")
)

copd_or_asthma <- picsure::buildClauseGroup(
  list(copd_filter, asthma_filter),
  operator = "OR"
)

filters <- picsure::buildClauseGroup(
  list(sex_filter, age_filter, copd_or_asthma),
  operator = "AND"
)
```

## Choosing output concepts

`buildQuery()` bundles a filter tree with the concept paths to return as
output columns:

```{r}
full_query <- picsure::buildQuery(
  phenotypicFilter = filters,
  includeConcepts  = c("\\phs000001\\pht000001\\phv00000020\\bmi\\")
)

# Include-only — return these concepts for every matching record, no filter:
bmi_only <- picsure::buildQuery(
  includeConcepts = c("\\phs000001\\pht000001\\phv00000020\\bmi\\")
)

# Filter only — a bare clause/clause-group runs without buildQuery():
count <- picsure::runQuery(bdc, filters, type = "count")
```

## Genomic filters

Genomic filters narrow a query to participants who carry variants matching
specific annotation values. They are available on authorized platforms
(BDC authorized, NHANES authorized) and are AND-combined with any
phenotypic filter.

Build a genomic filter with `buildGenomicFilter(key, values)`. `key` is a
`GenomicFilterKey` member (preferred) or the equivalent annotation string
(validated — an unknown string errors with the list of valid keys). `values`
is required — a character vector or `VariantFrequency` enum members.

```{r}
# Filter to BRCA1 or BRCA2 variants
gene_filter <- picsure::buildGenomicFilter(
  picsure::GenomicFilterKey$GENE_WITH_VARIANT,
  values = c("BRCA1", "BRCA2")
)

# Filter to rare variants using the VariantFrequency enum
rare_filter <- picsure::buildGenomicFilter(
  picsure::GenomicFilterKey$VARIANT_FREQUENCY_AS_TEXT,
  values = picsure::VariantFrequency$RARE
)
```

`Variant_severity` is a virtual key: `buildGenomicFilter` expands a
`VariantSeverity` bucket (`HIGH`, `MEDIUM`, `LOW`) into the matching
`Variant_consequence_calculated` values.

```{r}
# High-severity variants
severe_filter <- picsure::buildGenomicFilter(
  picsure::GenomicFilterKey$VARIANT_SEVERITY,
  values = picsure::VariantSeverity$HIGH
)
```

Pass a list of genomic filters to `buildQuery(genomicFilters = ...)` - they are
AND-combined. Genomic-only queries (no phenotypic filter) are allowed:

```{r}
# Genomic-only query: count participants carrying matching variants
genomic_query <- picsure::buildQuery(
  genomicFilters = list(gene_filter, rare_filter)
)
count <- picsure::runQuery(bdc, genomic_query, type = "count")

# Combined phenotypic + genomic query
combined_query <- picsure::buildQuery(
  phenotypicFilter = sex_filter,
  genomicFilters   = list(gene_filter)
)
```

`VariantFrequency` members (`RARE`, `COMMON`, `NOVEL`) are coerced to their
string value before the request is sent, so they are equivalent to passing
the string `"Rare"`, `"Common"`, or `"Novel"` directly.

## Editing an existing query

`removeSubQuery()` and `replaceClause()` let you edit a query tree
without rebuilding it from scratch. Both return a **new** handle — the
input is not mutated. Matching is structural: any nested clause or
clause-group that is equal by value to `target` is matched. They accept a
clause/clause-group or a Query (for a Query, the edit applies to its
phenotypic filter and the included concepts are preserved).

```{r}
# Drop the copd_or_asthma sub-tree:
without_lung <- picsure::removeSubQuery(full_query, copd_or_asthma)

# Swap "Yes" for a more specific COPD answer set:
copd_specific <- picsure::buildClause(
  "\\phs000001\\pht000002\\phv00000050\\copd\\",
  type = "FILTER",
  categories = list("Yes, severe", "Yes, very severe")
)
refined <- picsure::replaceClause(full_query, copd_filter, copd_specific)
```

`removeSubQuery()` drops `ClauseGroup`s that are emptied by a removal,
so you don't end up with orphan operators. It errors if the removal
would empty the entire tree — build a fresh query instead in that case.

## Save a query by name

On authorized deployments, persist a query to your user profile under
a display name with `saveQueryByName()`. It returns the PIC-SURE query
ID, which you can later pass to `loadQueryByID()` or `runQueryByID()`:

```{r}
qid <- picsure::saveQueryByName(bdc, full_query, "Cohort 2026-Q2")

# Re-point an existing name at a refined query:
qid <- picsure::saveQueryByName(bdc, refined, "Cohort 2026-Q2", overwrite = TRUE)
```

Allowed characters in `name`: letters, digits, spaces, and
``- _ \ / ? + = [ ] . ( ) : " '`` (max 255 chars; validated by the
Python adapter against the backend `NamedDataset` pattern). Not
supported on open-access platforms.

## Case-insensitive enums

Both `type` (clause type) and `operator` (group operator) accept
case-insensitive strings: `"filter"`, `"Filter"`, `"FILTER"` are all
equivalent.
