---
title: "Getting started with picsure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with picsure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

This vignette walks through installing `picsure`, connecting to a PIC-SURE
network, running a simple search, and executing a one-clause query.

## Install

```{r}
remotes::install_github("hms-dbmi/pic-sure-r-adapter-hpds@query_v3")
library(picsure)
```

The first call in any R session triggers reticulate to provision an
isolated Python environment. This takes a few seconds on first use; later
sessions reuse the cached environment.

## Connect

```{r}
bdc <- picsure::connect(
  platform = "BDC Authorized",
  token    = Sys.getenv("PICSURE_TOKEN")
)
```

Your token comes from the "User Profile" tab on the PIC-SURE web UI.
Valid platform names include `"Demo"`, `"BDC Open"`, and `"BDC Authorized"`.

## Search the dictionary

```{r}
picsure::search(bdc, "sex")
```

`search()` returns a data frame of matching variables. See
`vignette("search-and-facets")` for filtering by study or other facets.

## Build and run a query

```{r}
sex_filter <- picsure::createClause(
  "\\phs000001\\pht000001\\phv00000001\\sex\\",
  type       = "FILTER",
  categories = list("male")
)
query <- picsure::buildClauseGroup(list(sex_filter), operator = "AND")

count <- picsure::runQuery(bdc, query, type = "count")
count
```

See `vignette("building-queries")` for nested AND/OR trees and the other
clause types.
