---
title: "Getting started with picsure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with picsure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

This vignette walks through installing `picsure`, connecting to a PIC-SURE
network, running a simple search, and executing a one-clause query.

## Install

```{r}
remotes::install_github("hms-dbmi/pic-sure-r-adapter-hpds@v2.0.0")
library(picsure)
```

## Python environment

`picsure` declares its Python dependency via `reticulate::py_require()`,
and reticulate uses `uv` to materialize an isolated environment on the
first call that touches Python (typically `connect()`). There is no
manual `pip`, `conda`, or `virtualenv` step, and no separate
`install_picsure()` to invoke.

The first such call takes a few seconds; the environment is cached and
reused by later R sessions. To pay that cost up front — for example,
before a demo, or while still on VPN — just call `picsure::connect(...)`
early in the session.

Requires `reticulate >= 1.41` (already declared in `DESCRIPTION`); `uv`
is fetched by reticulate as needed.

## Connect

```{r}
bdc <- picsure::connect(
  platform = picsure::Platform$BDC_AUTHORIZED,
  token    = Sys.getenv("PICSURE_TOKEN")
)
```

Your token comes from the "User Profile" tab on the PIC-SURE web UI.
Pass a `Platform` member — e.g. `picsure::Platform$BDC_AUTHORIZED`,
`picsure::Platform$BDC_OPEN`, or `picsure::Platform$NHANES_OPEN` — or a
full URL string for an unlisted deployment. See `?picsure::Platform` for
the full set of members. A human-readable label string like
`"BDC Authorized"` is rejected.

## Search the dictionary

```{r}
picsure::searchDictionary(bdc, "sex")
```

`searchDictionary()` returns a data frame of matching variables. See
`vignette("search-and-facets")` for filtering by study or other facets.

## Build and run a query

```{r}
sex_filter <- picsure::buildClause(
  "\\phs000001\\pht000001\\phv00000001\\sex\\",
  type       = "FILTER",
  categories = list("male")
)
query <- picsure::buildClauseGroup(list(sex_filter), operator = "AND")

count <- picsure::runQuery(bdc, query, type = "count")
count
```

See `vignette("building-queries")` for nested AND/OR trees and the other
clause types.
