---
title: "Migrating from picsure v1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from picsure v1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

The 2.0.0 rewrite is a clean break from the 1.x adapter. Every v1 function
is gone; there are no deprecation shims. This page maps each v1 call to its
2.0.0 equivalent and walks through three common notebook patterns end-to-end.

## Function-by-function mapping

| v1 | v2 |
|---|---|
| `initializeSession(url, token)` | `connect(platform, token)` |
| `bdc.initializeSession(url, token)` | `connect(platform = Platform$BDC_AUTHORIZED, token)` |
| `bdc.setResource(session, "OPEN")` | `connect(platform = Platform$BDC_OPEN)` — no token needed |
| `bdc.searchPicsure(session, "sex")` | `searchDictionary(session, "sex")` |
| `bdc.getStudies(session)` | no dedicated call — the deployment's datasets are a facet category: `fs <- facets(session); addFacet(fs, "dataset_id", ...)` |
| `newQuery(session)` | no direct replacement — use `buildClause()` + `buildClauseGroup()` / `buildQuery()` |
| `addClause(q, path, "FILTER", min = 18)` | `buildClause(keys = path, type = "FILTER", min = 18)` |
| `addClause(q, path, "SELECT")` | `buildQuery(includeConcepts = path)` |
| `deleteClause(q, path)` | (removed — rebuild the tree, or edit with `removeSubQuery()` / `replaceClause()`) |
| `showQuery(q)` | (removed — queries are opaque Python handles) |
| `runQuery(q, "COUNT")` | `runQuery(session, query, type = "count")` — returns a `CountResult`, not a number (see below) |
| `runQuery(q, "DATA_FRAME")` | `runQuery(session, query, type = "participant")` |
| `getQueryByUUID(session, uuid)` | `loadQueryByID(session, query_id)` — rebuilds the saved query as a handle |
| `getResultByQueryUUID(session, uuid)` | `runQueryByID(session, query_id, type = "count")` — fetches and runs in one call |

New in 2.0.0, with no v1 equivalent:

| v2 | What it does |
|---|---|
| `saveQueryByName(session, query, name)` | submits the query and names it, returning its query ID |
| `buildGenomicFilter(key, values)` | a variant filter for `buildQuery(genomicFilters = ...)` |
| `searchGenomicValues(session, key)` | valid values for a genomic annotation key |
| `genomicConsequences()` | the offline variant-consequence vocabulary |
| `exportAsPFB(session, query, path)` | runs the query and writes a PFB file |
| `exportCSV(session, df, path)` / `exportTSV(...)` | write an already-materialized data frame; they do **not** re-run the query |
| `platforms()` | the human-readable labels of the known platforms (display only — `connect()` does not accept them) |

## `runQuery()` returns a result object, not a number

v1's `"COUNT"` handed back a bare number. `runQuery(session, query, type =
"count")` returns a `CountResult` with three fields: `$value` (the count, or
`NULL` when the deployment obfuscated a small cohort), `$margin`, and `$cap`.
Test for `NULL` before using it:

```{r}
count <- picsure::runQuery(bdc, query, type = "count")
if (is.null(count$value)) {
  message("cohort too small to report exactly (cap ", count$cap, ")")
} else {
  message(count$value, " participants")
}
```

The other result types are: `"participant"`, `"timestamp"`,
`"vcf_excerpt"`, and `"aggregate_vcf_excerpt"` return a data frame;
`"cross_count"` returns a named list of `CountResult`s keyed by concept path;
`"variant_count"` returns a `CountResult`; `"variant_list"` returns a
character vector.

## Worked example 1: simple count

**v1:**

```{r}
session <- picsure::bdc.initializeSession(
  url   = "https://picsure.biodatacatalyst.nhlbi.nih.gov/",
  token = my_token
)
q <- picsure::bdc.newQuery(session)
q <- picsure::bdc.addClause(
  q, "\\phs000001\\pht000001\\phv00000005\\age\\",
  "FILTER", min = 40
)
count <- picsure::runQuery(q, "COUNT")
```

**v2:**

```{r}
bdc <- picsure::connect(platform = picsure::Platform$BDC_AUTHORIZED, token = my_token)
age_filter <- picsure::buildClause(
  "\\phs000001\\pht000001\\phv00000005\\age\\",
  type = "FILTER", min = 40
)
query <- picsure::buildClauseGroup(list(age_filter), operator = "AND")
count <- picsure::runQuery(bdc, query, type = "count")
```

## Worked example 2: pull rows

**v1:**

```{r}
q <- picsure::bdc.newQuery(session)
q <- picsure::bdc.addClause(q, sex_path, "FILTER", categories = list("male"))
q <- picsure::bdc.addClause(q, bmi_path, "SELECT")
rows <- picsure::runQuery(q, "DATA_FRAME")
```

**v2:**

```{r}
sex_filter <- picsure::buildClause(sex_path, type = "FILTER", categories = list("male"))
query <- picsure::buildQuery(phenotypicFilter = sex_filter, includeConcepts = bmi_path)
rows  <- picsure::runQuery(bdc, query, type = "participant")
```

## Worked example 3: nested AND/OR

There's no v1 equivalent — v1 query construction was flat. In v2:

```{r}
copd    <- picsure::buildClause(copd_path,   type = "FILTER", categories = list("Yes"))
asthma  <- picsure::buildClause(asthma_path, type = "FILTER", categories = list("Yes"))
lung    <- picsure::buildClauseGroup(list(copd, asthma), operator = "OR")
query   <- picsure::buildClauseGroup(list(sex_filter, lung), operator = "AND")
```

## Error handling

v1 emitted raw HTTP / R errors from `httr`. The 2.0.0 adapter emits
`picsureError` conditions carrying researcher-facing messages. Catching that
one class still catches everything:

```{r}
tryCatch(
  picsure::runQuery(bdc, query),
  picsureError = function(e) {
    message("PIC-SURE error: ", conditionMessage(e))
  }
)
```

Subclasses let you tell the kinds of failure apart. `picsureAuthError` means
the server refused you — refresh the token; `picsureConnectionError` means no
usable response came back; `picsureQueryError` means the query itself was
rejected; `picsureValidationError` means this package rejected an argument
before any request was made. `?picsure::picsureError` has the full tree,
including which distinctions the currently pinned Python build can actually
report.

```{r}
tryCatch(
  picsure::runQuery(bdc, query),
  picsureAuthError    = function(e) message("Token problem: ", conditionMessage(e)),
  picsureQueryError   = function(e) message("Query rejected: ", conditionMessage(e)),
  picsureError        = function(e) message("PIC-SURE error: ", conditionMessage(e))
)
```

## Local and self-signed deployments

v1 took `httr` options directly. In 2.0.0 the HTTP client lives in Python, so
TLS verification and developer-mode instrumentation are set through R options
that `connect()` reads on every call:

```{r}
options(picsure.ssl_verify = FALSE)   # or a path to a CA bundle
options(picsure.dev_mode   = TRUE)
local <- picsure::connect(platform = "https://localhost", token = my_token)
```

Use the options rather than `Sys.setenv(PICSURE_SSL_VERIFY = ...)`: CPython
snapshots the environment when the interpreter starts, and reticulate runs
that interpreter inside the R process, so setting the environment variable
after the first call that provisions Python has no effect.

Note that a **custom URL** does not default to consent scoping the way
`Platform$BDC_AUTHORIZED` does. Pass `include_consents = TRUE` when
connecting by URL to a consent-gated deployment, or the session carries an
empty consent list and `searchDictionary()` returns every concept in the
index rather than the ones your consents cover.
