---
title: "Migrating from picsure v1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from picsure v1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

The 2.0.0 rewrite is a clean break from the 1.x adapter. Every v1 function
is gone; there are no deprecation shims. This page maps each v1 call to its
2.0.0 equivalent and walks through three common notebook patterns end-to-end.

## Function-by-function mapping

| v1 | v2 |
|---|---|
| `initializeSession(url, token)` | `connect(platform, token)` |
| `bdc.initializeSession(url, token)` | `connect(platform = Platform$BDC_AUTHORIZED, token)` |
| `bdc.setResource(session, "OPEN")` | `connect(platform = Platform$BDC_OPEN, ...)` |
| `bdc.searchPicsure(session, "sex")` | `searchDictionary(session, "sex")` |
| `bdc.getStudies(session)` | (removed — not yet re-exposed) |
| `newQuery(session)` | no direct replacement — use `buildClause()` + `buildClauseGroup()` / `buildQuery()` |
| `addClause(q, path, "FILTER", min = 18)` | `buildClause(keys = path, type = "FILTER", min = 18)` |
| `addClause(q, path, "SELECT")` | `buildQuery(includeConcepts = path)` |
| `deleteClause(q, path)` | (removed — rebuild the tree) |
| `showQuery(q)` | (removed — queries are opaque Python handles) |
| `runQuery(q, "COUNT")` | `runQuery(session, query, type = "count")` — returns a `CountResult` (use `$value` / `$cap`) |
| `runQuery(q, "DATA_FRAME")` | `runQuery(session, query, type = "participant")` |
| `getResultByQueryUUID(session, uuid)` | (removed — query UUID APIs not yet re-exposed) |
| `exportPFB(...)` (new) | `exportAsPFB(session, query, path)` |
| `exportCSV(...)` (new) | `df <- runQuery(...); exportCSV(session, df, path)` |
| `getQueryByUUID(session, uuid)` | (removed) |

## Worked example 1: simple count

**v1:**

```{r}
session <- picsure::bdc.initializeSession(
  url   = "https://picsure.biodatacatalyst.nhlbi.nih.gov/",
  token = my_token
)
q <- picsure::bdc.newQuery(session)
q <- picsure::bdc.addClause(
  q, "\\phs000001\\pht000001\\phv00000005\\age\\",
  "FILTER", min = 40
)
count <- picsure::runQuery(q, "COUNT")
```

**v2:**

```{r}
bdc <- picsure::connect(platform = picsure::Platform$BDC_AUTHORIZED, token = my_token)
age_filter <- picsure::buildClause(
  "\\phs000001\\pht000001\\phv00000005\\age\\",
  type = "FILTER", min = 40
)
query <- picsure::buildClauseGroup(list(age_filter), operator = "AND")
count <- picsure::runQuery(bdc, query, type = "count")
```

## Worked example 2: pull rows

**v1:**

```{r}
q <- picsure::bdc.newQuery(session)
q <- picsure::bdc.addClause(q, sex_path, "FILTER", categories = list("male"))
q <- picsure::bdc.addClause(q, bmi_path, "SELECT")
rows <- picsure::runQuery(q, "DATA_FRAME")
```

**v2:**

```{r}
sex_filter <- picsure::buildClause(sex_path, type = "FILTER", categories = list("male"))
query <- picsure::buildQuery(phenotypicFilter = sex_filter, includeConcepts = bmi_path)
rows  <- picsure::runQuery(bdc, query, type = "participant")
```

## Worked example 3: nested AND/OR

There's no v1 equivalent — v1 query construction was flat. In v2:

```{r}
copd    <- picsure::buildClause(copd_path,   type = "FILTER", categories = list("Yes"))
asthma  <- picsure::buildClause(asthma_path, type = "FILTER", categories = list("Yes"))
lung    <- picsure::buildClauseGroup(list(copd, asthma), operator = "OR")
query   <- picsure::buildClauseGroup(list(sex_filter, lung), operator = "AND")
```

## Error handling

v1 emitted raw HTTP / R errors from `httr`. The 2.0.0 adapter emits `picsureError` with
Python-crafted, researcher-facing messages. Catch the class if you need to:

```{r}
tryCatch(
  picsure::runQuery(bdc, query),
  picsureError = function(e) {
    message("PIC-SURE error: ", conditionMessage(e))
  }
)
```
