---
title: "Running and exporting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running and exporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

Given a query built with `buildClause()` / `buildClauseGroup()` (and
optionally `buildQuery()`), you can ask the server for:

- a participant count (possibly obfuscated for small cohorts),
- a participant-level data frame,
- a timestamped longitudinal data frame,
- a PFB file written directly to disk.

CSV and TSV output work differently: they serialize a DataFrame you
already have, so the flow is "runQuery first, then write".

## Run for a count

```{r}
count <- picsure::runQuery(bdc, full_query, type = "count")
```

`count` is a `CountResult` object with three fields:

- `count$value` — the exact count, or `NULL` for obfuscated small cohorts.
- `count$margin` — the server-side margin for the count (when reported).
- `count$cap` — the upper bound used when a count is obfuscated.

```{r}
if (!is.null(count$value)) {
  cat(count$value, "participants\n")
} else {
  cat("fewer than", count$cap, "participants\n")
}
```

## Cross counts

Pass `type = "cross_count"` to receive a dict-like object keyed by concept
path, each value a `CountResult`.

```{r}
cc <- picsure::runQuery(bdc, full_query, type = "cross_count")
```

## Pull participant rows

```{r}
participants <- picsure::runQuery(bdc, full_query, type = "participant")
nrow(participants)
head(participants)
```

## Timestamped data

```{r}
timeseries <- picsure::runQuery(bdc, full_query, type = "timestamp")
```

## Export to disk

`exportAsPFB()` runs the query and writes the PFB file in a single call:

```{r}
picsure::exportAsPFB(bdc, full_query, "~/cohort.pfb")
```

PFB export requires the Python `picsure[pfb]` optional dependency. If the
Python env was provisioned without it, `exportAsPFB()` raises a
`picsureError` pointing you at the install hint.

`exportCSV()` and `exportTSV()` serialize an R `data.frame` — the typical
flow is to run the query first and then write the result:

```{r}
df <- picsure::runQuery(bdc, full_query, type = "participant")
picsure::exportCSV(bdc, df, "~/cohort.csv")
picsure::exportTSV(bdc, df, "~/cohort.tsv")
```

All three export functions return the path invisibly, so you can chain:

```{r}
pfb_path <- picsure::exportAsPFB(bdc, full_query, tempfile(fileext = ".pfb"))
```

## Save a query by name

On authorized deployments, `saveQueryByName()` persists a query to
the user's profile under a display name. It submits the query to
PIC-SURE and associates the name with it server-side, returning the
PIC-SURE query ID:

```{r}
qid <- picsure::saveQueryByName(bdc, full_query, "Cohort 2026-Q2")
```

The returned UUID is the same one you'd pass to `loadQueryByID()` or
`runQueryByID()`:

```{r}
restored <- picsure::loadQueryByID(bdc, qid)
df       <- picsure::runQueryByID(bdc, qid, type = "participant")
```

By default, saving a second time under the same name is rejected. To
repoint the existing record at a freshly-built query, pass
`overwrite = TRUE`:

```{r}
qid <- picsure::saveQueryByName(bdc, refined, "Cohort 2026-Q2", overwrite = TRUE)
```

`name` must be a non-empty character scalar; allowed characters are
letters, digits, spaces, and ``- _ \ / ? + = [ ] . ( ) : " '`` (max
255). `saveQueryByName()` is not supported on open-access platforms —
the `/dataset/named/` endpoint requires an authenticated principal.

## Reload a previously-saved query

If you have the UUID of a previously-saved query, `loadQueryByID()` fetches
the saved body from the backend and rebuilds it as a Clause / ClauseGroup
that drops straight back into `runQuery()` or any of the export functions:

```{r}
previous <- picsure::loadQueryByID(bdc, "11111111-2222-3333-4444-555555555555")
count <- picsure::runQuery(bdc, previous, type = "count")
```

The loaded handle is a clause/clause-group (or a Query, when the saved
query selected output concepts) — the same shapes the builders produce —
so a bare clause/group can also be nested inside a new group to refine
the cohort.
