---
title: "Running and exporting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running and exporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

Given a query built with `createClause()` + `buildClauseGroup()`, you can
ask the server for:

- a participant count (possibly obfuscated for small cohorts),
- a participant-level data frame,
- a timestamped longitudinal data frame,
- a PFB file written directly to disk.

CSV and TSV output work differently: they serialize a DataFrame you
already have, so the flow is "runQuery first, then write".

## Run for a count

```{r}
count <- picsure::runQuery(bdc, full_query, type = "count")
```

`count` is a `CountResult` object with three fields:

- `count$value` — the exact count, or `NULL` for obfuscated small cohorts.
- `count$margin` — the server-side margin for the count (when reported).
- `count$cap` — the upper bound used when a count is obfuscated.

```{r}
if (!is.null(count$value)) {
  cat(count$value, "participants\n")
} else {
  cat("fewer than", count$cap, "participants\n")
}
```

## Cross counts

Pass `type = "cross_count"` to receive a dict-like object keyed by concept
path, each value a `CountResult`.

```{r}
cc <- picsure::runQuery(bdc, full_query, type = "cross_count")
```

## Pull participant rows

```{r}
participants <- picsure::runQuery(bdc, full_query, type = "participant")
nrow(participants)
head(participants)
```

## Timestamped data

```{r}
timeseries <- picsure::runQuery(bdc, full_query, type = "timestamp")
```

## Export to disk

`exportPFB()` runs the query and writes the PFB file in a single call:

```{r}
picsure::exportPFB(bdc, full_query, "~/cohort.pfb")
```

PFB export requires the Python `picsure[pfb]` optional dependency. If the
Python env was provisioned without it, `exportPFB()` raises a
`picsure_error` pointing you at the install hint.

`exportCSV()` and `exportTSV()` serialize an R `data.frame` — the typical
flow is to run the query first and then write the result:

```{r}
df <- picsure::runQuery(bdc, full_query, type = "participant")
picsure::exportCSV(bdc, df, "~/cohort.csv")
picsure::exportTSV(bdc, df, "~/cohort.tsv")
```

All three export functions return the path invisibly, so you can chain:

```{r}
pfb_path <- picsure::exportPFB(bdc, full_query, tempfile(fileext = ".pfb"))
```
