---
title: "Running and exporting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running and exporting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

Given a query built with `createClause()` + `buildClauseGroup()`, you can
ask the server for:

- a participant count,
- a participant-level data frame,
- a timestamped longitudinal data frame,
- a PFB, CSV, or TSV file on disk.

## Run for a count

```{r}
count <- picsure::runQuery(bdc, full_query, type = "count")
count
```

## Pull participant rows

```{r}
participants <- picsure::runQuery(bdc, full_query, type = "participant")
nrow(participants)
head(participants)
```

## Timestamped data

```{r}
timeseries <- picsure::runQuery(bdc, full_query, type = "timestamp")
```

## Export to disk

```{r}
picsure::exportCSV(bdc, full_query, "~/cohort.csv")
picsure::exportTSV(bdc, full_query, "~/cohort.tsv")
picsure::exportPFB(bdc, full_query, "~/cohort.pfb")
```

PFB export requires the Python `picsure[pfb]` optional dependency. If the
Python env was provisioned without it, `exportPFB()` raises a
`picsureError` pointing you at the install hint.

All three export functions return the path invisibly, so you can chain:

```{r}
pfb_path <- picsure::exportPFB(bdc, full_query, tempfile(fileext = ".pfb"))
```
