---
title: "Search and facets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Search and facets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

`picsure::searchDictionary()` returns matching dictionary entries as a data frame.
`picsure::facets()` narrows a search to a subset — by study, by consent
group, etc.

## Plain search

```{r}
library(picsure)
bdc <- picsure::connect(platform = picsure::Platform$BDC_AUTHORIZED, token = Sys.getenv("PICSURE_TOKEN"))

results <- picsure::searchDictionary(bdc, "sex")
head(results)
```

## Build a FacetSet

```{r}
fs <- picsure::facets(bdc)
fs <- picsure::addFacet(fs, "study_ids", "phs000007")
```

`addFacet()` accepts a vector of values to add several at once:

```{r}
fs <- picsure::addFacet(fs, "study_ids", c("phs000200", "phs000286"))
```

Remove with `removeFacet(fs, key, value)`. Both functions mutate the
underlying Python FacetSet and return it invisibly for chaining.

## Search with facets

```{r}
picsure::searchDictionary(bdc, "sex", facets = fs)
```

The server restricts the search to variables inside the facets.

## Lighter payloads

Pass `include_values = FALSE` to omit variable values from the response —
useful when you only need variable names and descriptions:

```{r}
picsure::searchDictionary(bdc, "age", include_values = FALSE)
```

The search term is always matched server-side; you receive every matching
row. There is no `limit` or `offset` — narrow with a more specific term
or a facet filter if a search returns too many results.
